Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-09-25 draftfinalreport_v4.Rnw
adds bit about not covering entire MMR program to discussion re B/C rations
fixup to work with JM's table

vacc_dhb <- read.csv("tables/dhb_vacc.csv")
totals   <- vacc_dhb[nrow(vacc_dhb), ,drop=FALSE]
vacc_reqd_percent    <- totals$PC * 100
vacc_reqd_dhb_min    <- min(vacc_dhb$PC) * 100
vacc_reqd_dhb_max    <- max(vacc_dhb$PC) * 100
vacc_reqd_auckland   <- format(min(floor((vacc_dhb %>% filter(DHB == "Auckland" | DHB == "CountiesManakau" | DHB == "Waitemata") %>% select(Vaccination)) / 1000)) * 1000, big.mark=",")
vacc_reqd_count      <- format(totals$Vaccination, big.mark=",")
naive_percent        <- round(totals$Naive / totals$Population * 100)
break_even_vacc_cost <- format(74, big.mark=",") # vaccine costs such that benefits == cost (from .xlsx)
break_even_vacc_cost50 <- format(35.70, big.mark=",") # vaccine costs such that benefits == cost (from .xlsx)
@

@@

change<-read.csv("change.csv",header=T)
costtable1<-read.csv("costtable1.csv",header=T)
costtable2<-read.csv("costtable2v2.csv",header=T)
costtable3<-read.csv("costtable3.csv",header=T)
costtable4<-read.csv("costtable4v2.csv",header=T)
costtable5<-read.csv("costtable5v2.csv",header=T)
costtable20<-read.csv("tables/cost_benefit_20.csv",header=T)
costtable50<-read.csv("tables/cost_benefit_50.csv",header=T)
costtable74<-read.csv("tables/cost_benefit_74.csv",header=T)
costtable20_no<-read.csv("tables/cost_benefit_no_20.csv",header=T)
costtable50_no<-read.csv("tables/cost_benefit_no_50.csv",header=T)
costtable_50pc<-read.csv("tables/cost_benefit_50pc.csv",header=T)

@@

\begin{sidewaystable}\footnotesize
<<results=tex,echo=FALSE>>=
colnames(costtable_50pc)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hosp saved", "Contacts","Costs saved", "Outbreak", "OB costs", "B/C","NPV")
costtable_50pc[,c(2:10,12)] <- round(costtable_50pc[,c(2:10,12)])
costtable_50pc[,11] <- round(costtable_50pc[,11],2)
latex(costtable_50pc, file="", table.env=FALSE,rowname=NULL)
@
\caption{Benefit--cost analyses with \$35.74 dollars per immunised person assuming the cases prevented through additional vaccination each year are equal to the mean annual number of cases per DHB since 2009 (see \autoref{table:vacp}). This is the break-even cost per immunised person (B/C = 1) nationally if 50\% of the currently na\"{\i}ve population were vaccinated. \textit{Vacc} is numbers to immunise if 50\% of the currently na\"{\i}ve population were vaccinated (see \autoref{table:vacp}); \textit{Vacc costs} is cost for the catch up immunisation programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion (\$2,624.57 per case); \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted median outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the median outbreak sizes are what is expected, while the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introductions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio; \textit{NPV} is the net present value, defined as the sum of the present values of costs and benefits of the vaccination programme.}
\label{table:cost50pc}
\end{sidewaystable}

@@

The last benefit--cost analysis scenario were performed was one in which more than the minimum proportion of na\"{\i}ve were immunised. In this scenario we did not use the \Sexpr{vacc_reqd_percent}\% of currently na\"{\i}ve population necessary to reach adequate national immunity levels (\autoref{table:coh_vac}), but included additional vaccination so that 50$\%$ of the currently na\"{\i}ve population was immunised, \autoref{table:cost50pc}. In this scenario the additional vaccination reduces the B/C ratio, and the additional vaccines for the extra immunisation mean that the cost per immunised person before the programme becomes costly reduces to \$35.74  per immunised person \label{table:cost50pc}.

@@

Our estimates of the costs of the measles outbreaks in New Zealand suggest that measles management costs per case in New Zealand is high. We used the mean values per case in the absence of alternative data and used all available data. The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up immunisation schemes will be financially beneficial for New Zealand (see \autoref{table:cost20} and \autoref{table:cost50}). Additional vaccination beyond the \Sexpr{vacc_reqd_percent}\% of currently na\"{\i}ve population we estimate are required to be vaccinated for $R_0 < 1$ decrease the cost effectiveness of an immunisation program. This is because large outbreaks are avoided once $R_0 < 1$. In the scenario we modelled with 50\% of the currently na\"{\i}ve population immunised, the programme become costly at \$35.74  per immunised person \label{table:cost50pc}. There may be other benefits we have not accounted for in our analyses for additional vaccination. These additional benefits may include reducing the possibility of spatial heterogeneity and future benefits beyond the ten year period we modelled.

@@

Our estimates for the benefit--cost ratio of catch up immunisation are similar to those in other countries. Estimates in Korea suggest catch up immunisation schemes have a benefit--cost ratio of just over one \citep{bae13}. If we presume the $R_0$ estimated for measles in New Zealand is representative and following \Sexpr{vacc_reqd_percent}\% immunisation of the na\"{\i}ve population, measles will be unable to persist ($R_v$ < 1), our B/C ratio is greater than 1. At the national level our B/C ratio is > 1 for either \$50 or \$20 effective immunisation costs. However, the B/C ratios do vary substantially by DHB, with those with smaller populations least cost effective for immunisation programmes. However, the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see \autoref{table:zhou2}) \citep{zhou4}. As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis. Our results are relatively insensitive to changes in immunisation cost, requiring high costs (>\$\Sexpr{break_even_vacc_cost} per immunised person at the national level) before immunisation is predicted to not be beneficial. Similarly, longer discounting periods or 5\% discounting did not affect the B/C ratio substantially (not shown). A spreadsheet is provided for the Ministry to work with.

>>>>

vacc_dhb <- read.csv("tables/dhb_vacc.csv")
totals   <- vacc_dhb[nrow(vacc_dhb), ,drop=FALSE]
vacc_reqd_percent    <- totals$PC * 100
vacc_reqd_dhb_min    <- min(vacc_dhb$PC) * 100
vacc_reqd_dhb_max    <- max(vacc_dhb$PC) * 100
vacc_reqd_auckland   <- format(min(floor((vacc_dhb %>% filter(DHB == "Auckland" | DHB == "CountiesManakau" | DHB == "Waitemata") %>% select(Vaccination)) / 1000)) * 1000, big.mark=",")
vacc_reqd_count      <- format(totals$Vaccination, big.mark=",")
naive_percent        <- round(totals$Naive / totals$Population * 100)
break_even_vacc_cost <- format(74, big.mark=",") # vaccine costs such that benefits == cost (from .xlsx)
break_even_vacc_cost50 <- sprintf("%0.2f", 35.70) # vaccine costs such that benefits == cost (from .xlsx)
@

@@

change<-read.csv("change.csv",header=T)
costtable1<-read.csv("costtable1.csv",header=T)
costtable2<-read.csv("costtable2v2.csv",header=T)
costtable3<-read.csv("costtable3.csv",header=T)
costtable4<-read.csv("costtable4v2.csv",header=T)
costtable5<-read.csv("costtable5v2.csv",header=T)
costtable20<-read.csv("tables/cost_benefit_20.csv",header=T)
costtable50<-read.csv("tables/cost_benefit_50.csv",header=T)
costtable74<-read.csv("tables/cost_benefit_74.csv",header=T)
costtable20_no<-read.csv("tables/cost_benefit_no_20.csv",header=T)
costtable50_no<-read.csv("tables/cost_benefit_no_50.csv",header=T)
costtable_50pc<-read.csv("tables/cost_benefit_50_percent_vacc_35.csv",header=T)

@@

\begin{sidewaystable}\footnotesize
<<results=tex,echo=FALSE>>=
colnames(costtable_50pc)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hosp saved", "Contacts","Costs saved", "Outbreak", "OB costs", "B/C","NPV")
costtable_50pc[,c(2:10,12)] <- round(costtable_50pc[,c(2:10,12)])
costtable_50pc[,11] <- round(costtable_50pc[,11],2)
latex(costtable_50pc, file="", table.env=FALSE,rowname=NULL)
@
\caption{Benefit--cost analyses with \$\Sexpr{break_even_vacc_cost50} dollars per immunised person assuming the cases prevented through additional vaccination each year are equal to the mean annual number of cases per DHB since 2009 (see \autoref{table:vacp}). This is the break-even cost per immunised person (B/C = 1) nationally if 50\% of the currently na\"{\i}ve population were vaccinated. \textit{Vacc} is numbers to immunise if 50\% of the currently na\"{\i}ve population were vaccinated (see \autoref{table:vacp}); \textit{Vacc costs} is cost for the catch up immunisation programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion (\$2,624.57 per case); \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted median outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the median outbreak sizes are what is expected, while the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introductions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio; \textit{NPV} is the net present value, defined as the sum of the present values of costs and benefits of the vaccination programme.}
\label{table:cost50pc}
\end{sidewaystable}

@@

The last benefit--cost analysis scenario were performed was one in which more than the minimum proportion of na\"{\i}ve were immunised. In this scenario we did not use the \Sexpr{vacc_reqd_percent}\% of currently na\"{\i}ve population necessary to reach adequate national immunity levels (\autoref{table:coh_vac}), but included additional vaccination so that 50$\%$ of the currently na\"{\i}ve population was immunised, \autoref{table:cost50pc}. In this scenario the additional vaccination reduces the B/C ratio, and the additional vaccines for the extra immunisation mean that the cost per immunised person before the programme becomes costly reduces to \$\Sexpr{break_even_vacc_cost50}  per immunised person \label{table:cost50pc}.

@@

Our estimates of the costs of the measles outbreaks in New Zealand suggest that measles management costs per case in New Zealand is high. We used the mean values per case in the absence of alternative data and used all available data. The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up immunisation schemes will be financially beneficial for New Zealand (see \autoref{table:cost20} and \autoref{table:cost50}). Additional vaccination beyond the \Sexpr{vacc_reqd_percent}\% of currently na\"{\i}ve population we estimate are required to be vaccinated for $R_0 < 1$ decrease the cost effectiveness of an immunisation program. This is because large outbreaks are avoided once $R_0 < 1$. In the scenario we modelled with 50\% of the currently na\"{\i}ve population immunised, the programme become costly at \$\Sexpr{break_even_vacc_cost50} per immunised person \label{table:cost50pc}. There may be other benefits we have not accounted for in our analyses for additional vaccination. These additional benefits may include reducing the possibility of spatial heterogeneity and future benefits beyond the ten year period we modelled.

@@

Our estimates for the benefit--cost ratio of catch up immunisation are similar to those in other countries. Estimates in Korea suggest catch up immunisation schemes have a benefit--cost ratio of just over one \citep{bae13}. If we presume the $R_0$ estimated for measles in New Zealand is representative and following \Sexpr{vacc_reqd_percent}\% immunisation of the na\"{\i}ve population, measles will be unable to persist ($R_v$ < 1), our B/C ratio is greater than 1. At the national level our B/C ratio is > 1 for either \$50 or \$20 effective immunisation costs. However, the B/C ratios do vary substantially by DHB, with those with smaller populations least cost effective for immunisation programmes. However, the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see \autoref{table:zhou2}) \citep{zhou4}. As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis. We did not include the B/C ratio of the complete MMR immunisation programme, though presumably if this was included our B/C ratio may  approached the ratios reported in those analyses that do \citep{zhou4}. Our results are relatively insensitive to changes in immunisation cost, requiring high costs (>\$\Sexpr{break_even_vacc_cost} per immunised person at the national level) before immunisation is predicted to not be beneficial. Similarly, longer discounting periods or 5\% discounting did not affect the B/C ratio substantially (not shown). A spreadsheet is provided for the Ministry to work with.
